---
title: "Corpus"
---

```{r setup}
#| include: false
source("setup.R")
```

## The Two Playlists

The corpus consists of 861 tracks drawn from two Spotify playlists, each representing one end of the EDM energy spectrum. Both playlists are embedded below for listening reference.

### Groovy Minimal House

<span class="genre-tag minimal">536 tracks — hypnotic basslines, groove-driven repetition</span>

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/2Alz8zsZFJLzDthh37proC?utm_source=generator" width="100%" height="352" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" loading="lazy"></iframe>

### Club Heartbroken

<span class="genre-tag hard">325 tracks — aggressive kicks, maximal compression</span>

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/5uOkatltA5XM6mN4HHeIT6?utm_source=generator" width="100%" height="352" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" loading="lazy"></iframe>

## Full Corpus Explorer

The table below contains all 861 tracks with their Spotify audio features. Use the column filters to search by genre, artist, or feature range.

```{r}
corpus |> select(Genre,`Track Name`,`Artist Name(s)`,Tempo,Energy,Danceability,Valence,Loudness,Popularity) |>
  mutate(Tempo=round(Tempo,1),Energy=round(Energy,3),Danceability=round(Danceability,3),Valence=round(Valence,3),Loudness=round(Loudness,1)) |>
  datatable(filter="top",options=list(pageLength=15,scrollX=T,dom='frtip',columnDefs=list(list(className='dt-center',targets=3:8))),
    class="table table-striped table-hover table-sm",rownames=F)
```
