---
title: "Week 12: Classification & Clustering"
---

```{r setup}
#| include: false
source("setup.R")
library(tidymodels)
library(ggdendro)
corpus_clean<-corpus|>drop_na(Danceability,Energy,Loudness,Speechiness,Acousticness,Instrumentalness,Liveness,Valence,Tempo,`Duration (ms)`)|>mutate(Genre=factor(Genre))
class_recipe<-recipe(Genre~Danceability+Energy+Loudness+Speechiness+Acousticness+Instrumentalness+Liveness+Valence+Tempo+`Duration (ms)`,data=corpus_clean)|>step_center(all_predictors())|>step_scale(all_predictors())
set.seed(42);corpus_cv<-vfold_cv(corpus_clean,v=10)
forest_model<-rand_forest()|>set_mode("classification")|>set_engine("ranger",importance="impurity")
forest_fit<-workflow()|>add_recipe(class_recipe)|>add_model(forest_model)|>fit_resamples(corpus_cv,control=control_resamples(save_pred=T))
accuracy_val<-forest_fit|>collect_metrics()|>filter(.metric=="accuracy")|>pull(mean)|>round(3)
```

## Classification and Clustering

This section tests whether Spotify's audio features can reliably distinguish Minimal Tech House from Hard House using supervised classification (random forest) and unsupervised clustering (hierarchical). Together, these methods provide both a performance benchmark and an independent validation of the genre distinction.

## Confusion Matrix: Random Forest (10-fold CV)

A random forest classifier is trained using `tidymodels` with 10-fold stratified cross-validation. The confusion matrix heatmap shows how many tracks from each genre are correctly or incorrectly classified.

```{r}
#| code-summary: "Confusion matrix: random forest 10-fold CV"
#| fig-width: 7
#| fig-height: 5
#| out-width: "100%"
forest_fit|>collect_predictions()|>conf_mat(truth=Genre,estimate=.pred_class)|>autoplot(type="heatmap")+
  scale_fill_viridis_c(option="plasma")+labs(title="Confusion Matrix: Random Forest (10-fold CV)")+
  theme_freq(13)+theme(panel.grid=element_blank(),legend.position="none")
```

The random forest classifier achieves a 10-fold cross-validated accuracy of `r accuracy_val`, which shows that Minimal Tech House and Hard House are computationally distinguishable with high reliability. Looking at the confusion matrix in detail: out of 325 Hard House tracks, 321 are correctly classified and only 4 are misclassified as Minimal Tech House. Out of 535 Minimal Tech House tracks, none or almost none are misclassified as Hard House. The few misclassifications likely correspond to boundary tracks that sit in the overlap region visible in the Week 8 scatter plot, where tempo and energy values converge between the two genres. The near-zero false positive rate for Hard House is particularly notable, suggesting that the features that define Hard House (high tempo, high loudness, high energy) are highly distinctive and rarely found in Minimal Tech House tracks.

## Feature Importance

Feature importance is extracted from the fitted random forest using the `vip` package. The importance score reflects how much each feature reduces impurity (Gini) across all decision trees. Higher values indicate features that contribute more to separating the genres.

```{r}
#| code-summary: "Feature importance: impurity-based variable importance"
#| fig-width: 10
#| fig-height: 5
#| out-width: "100%"
importance_fit<-workflow()|>add_recipe(class_recipe)|>add_model(forest_model)|>fit(corpus_clean)
importance_fit|>extract_fit_parsnip()|>vip::vi()|>mutate(Variable=fct_reorder(Variable,Importance))|>
  ggplot(aes(x=Importance,y=Variable,fill=Importance))+geom_col()+scale_fill_viridis_c(guide="none",option="plasma")+
  theme_freq(12)+labs(title="Feature Importance: Random Forest",subtitle="Which Spotify features best separate the two genres?",x="Importance (impurity reduction)",y=NULL)
```

The feature importance ranking shows which audio features the classifier relies on most heavily. Tempo dominates, which aligns with the near-complete separation visible in the Week 11 histogram. Danceability, Energy, and Loudness follow closely, all features that directly relate to the physical experience of each genre. Features like Speechiness and Liveness contribute almost nothing to the classification, which makes sense for two purely electronic, instrumental genres where neither speech nor live performance characteristics play a role.

## Reduced Model: Top 5 Features

To test whether the full 10-feature model is necessary, a reduced random forest is trained using only the top 5 features from the importance ranking. This tests the robustness and parsimony of the genre distinction.

```{r}
#| code-summary: "Reduced model: train with top 5 features only"
reduced_recipe<-recipe(Genre~Tempo+Loudness+Energy+Danceability+Valence,data=corpus_clean)|>step_center(all_predictors())|>step_scale(all_predictors())
reduced_fit<-workflow()|>add_recipe(reduced_recipe)|>add_model(forest_model)|>fit_resamples(corpus_cv,control=control_resamples(save_pred=T))
reduced_accuracy<-reduced_fit|>collect_metrics()|>filter(.metric=="accuracy")|>pull(mean)|>round(3)
```

```{r}
#| code-summary: "Confusion matrix: reduced model"
#| fig-width: 7
#| fig-height: 5
#| out-width: "100%"
reduced_fit|>collect_predictions()|>conf_mat(truth=Genre,estimate=.pred_class)|>autoplot(type="heatmap")+
  scale_fill_viridis_c(option="plasma")+labs(title="Reduced Model (Top 5 Features)")+
  theme_freq(13)+theme(panel.grid=element_blank(),legend.position="none")
```

The reduced model, using only Tempo, Loudness, Energy, Danceability, and Valence, achieves an accuracy of **`r reduced_accuracy`** compared to **`r accuracy_val`** with all ten features. The negligible difference confirms that the genre distinction is primarily encoded in a small set of features tied to tempo, loudness, and energy. The five removed features (Speechiness, Liveness, Instrumentalness, Acousticness, Duration) add almost no discriminatory power for this particular genre pair.

## Hierarchical Clustering: Dendrogram

As an unsupervised complement to the supervised classifier, hierarchical clustering is applied to a random sample of 60 tracks (30 per genre). Features are standardised, Euclidean distances are computed, and average-linkage clustering groups similar tracks together. The dendrogram is coloured by genre to assess whether the unsupervised structure aligns with the known labels.

```{r}
#| code-summary: "Dendrogram: hierarchical clustering of 60 tracks"
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
set.seed(123)
cs<-corpus_clean|>group_by(Genre)|>slice_sample(n=30)|>ungroup()|>mutate(label=paste0(str_trunc(`Track Name`,20)," [",if_else(Genre=="Hard House","HH","MT"),"]"))
cf<-cs|>select(label,Danceability,Energy,Loudness,Speechiness,Acousticness,Instrumentalness,Liveness,Valence,Tempo,`Duration (ms)`)|>column_to_rownames("label")
cl<-dist(scale(cf),method="euclidean")|>hclust(method="average");dd<-dendro_data(cl)
ldf<-dd$labels|>mutate(genre=if_else(str_detect(label,"\\[HH\\]"),"Hard House","Minimal Tech House"),col=if_else(genre=="Hard House","#ff2d55","#00d4ff"))
ggplot()+geom_segment(data=dd$segments,aes(x=x,y=y,xend=xend,yend=yend),colour="grey40",linewidth=0.4)+
  geom_text(data=ldf,aes(x=x,y=y-0.2,label=label,colour=col),hjust=1,size=2.3,angle=90)+
  scale_colour_identity()+scale_y_continuous(expand=expansion(mult=c(0.35,0.05)))+
  theme_freq(11)+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(),panel.grid=element_blank())+
  labs(title="Hierarchical Clustering: 60-Track Sample",subtitle="Average linkage | Euclidean distance | red = Hard House, cyan = Minimal Tech House",x=NULL,y="Distance")
```

The dendrogram provides an unsupervised validation of the supervised classification results. Using average-linkage hierarchical clustering on standardised features, the 60-track sample splits into two clearly distinct branches that align closely with the genre labels. Hard House tracks (red) group together on one side, while Minimal Tech House tracks (cyan) form a separate cluster. The fact that unsupervised clustering recovers the same genre boundary that the supervised classifier learned, without access to any labels, provides strong converging evidence that the distinction between these genres reflects genuine, measurable differences in audio features rather than merely a labelling convention.
