---
title: "Week 10: Tonal Centre"
---

```{r setup}
#| include: false
source("setup.R")
```

## Tonal Centre Analysis

This section explores tonal structure through key estimation. A global key profile shows which pitch classes dominate across an entire track, while a keygram tracks how key estimates change over time. These methods are especially informative for EDM, where harmonic palettes are deliberately narrow and heavily reinforced.

## Global Key Profile: Wood

The key profile is computed by averaging the NNLS Chroma intensities across the full track for each of the 12 pitch classes. The resulting bar chart acts as a tonal fingerprint, revealing which notes dominate the harmonic content.

```{r}
#| code-summary: "Key profile: average chroma intensity per pitch class"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"

# Load chroma data exported from Sonic Visualiser (NNLS Chroma plugin)
wood_data <- read_csv("wood.csv", show_col_types = FALSE)

# Average each pitch class intensity across the full track duration
wood_avg <- wood_data |>
  select(-TIME) |>
  summarise(across(everything(), mean)) |>
  pivot_longer(everything(), names_to = "Pitch_Class", values_to = "Intensity") |>
  mutate(Pitch_Class = factor(Pitch_Class,
    levels = c("C","C#","D","Eb","E","F","F#","G","Ab","A","Bb","B")))

# Bar chart showing which pitch classes dominate
ggplot(wood_avg, aes(x = Pitch_Class, y = Intensity, fill = Intensity)) +
  geom_col(width = 0.7) +
  scale_fill_viridis_c(guide = "none", option = "plasma") +
  theme_freq(13) +
  labs(
    title = "Overall Key Profile: 'Wood'",
    subtitle = "Average pitch-class intensity across the full track",
    x = "Pitch Class", y = "Average Intensity"
  )
```

The bar chart shows the average intensity of each pitch class across the entire duration of *Wood*. A small number of pitch classes dominate clearly, while others are nearly absent. The strongest peaks appear on F, C, and G, with secondary emphasis on Bb and Eb. This pattern is consistent with an F minor or Ab major tonality (which share the same pitch class set), though the absence of strong thirds makes it difficult to distinguish major from minor with certainty. In EDM, where harmonic material is deliberately narrow and heavily reinforced through repetition, this profile acts as a compact fingerprint of tonal emphasis. The pronounced peaks connect directly to the loop-based patterns visible in the Week 9 chroma analysis: the same few notes that form the harmonic loop are the ones that accumulate the highest average intensity over time.

## Keygram: Wood

The keygram computes a Euclidean distance between the chroma profile of each time window and the Krumhansl-Kessler key templates for all 24 major and minor keys. Darker cells indicate a closer match. This method reveals whether and when key changes occur.

```{r}
#| code-summary: "Keygram: Krumhansl-Kessler key estimation over time"
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
circshift<-function(v,n){if(n==0)v else c(tail(v,n),head(v,-n))}
major_key<-c(6.35,2.23,3.48,2.33,4.38,4.09,2.52,5.19,2.39,3.66,2.29,2.88)
minor_key<-c(6.33,2.68,3.52,5.38,2.60,3.53,2.54,4.75,3.98,2.69,3.34,3.17)
key_templates<-tribble(~name,~template,"Gb:maj",circshift(major_key,6),"Bb:min",circshift(minor_key,10),"Db:maj",circshift(major_key,1),"F:min",circshift(minor_key,5),"Ab:maj",circshift(major_key,8),"C:min",circshift(minor_key,0),"Eb:maj",circshift(major_key,3),"G:min",circshift(minor_key,7),"Bb:maj",circshift(major_key,10),"D:min",circshift(minor_key,2),"F:maj",circshift(major_key,5),"A:min",circshift(minor_key,9),"C:maj",circshift(major_key,0),"E:min",circshift(minor_key,4),"G:maj",circshift(major_key,7),"B:min",circshift(minor_key,11),"D:maj",circshift(major_key,2),"F#:min",circshift(minor_key,6),"A:maj",circshift(major_key,9),"C#:min",circshift(minor_key,1),"E:maj",circshift(major_key,4),"G#:min",circshift(minor_key,8),"B:maj",circshift(major_key,11),"D#:min",circshift(minor_key,3))
pitch_cols<-c("C","C#","D","Eb","E","F","F#","G","Ab","A","Bb","B")
tm<-do.call(rbind,key_templates$template);tm<-tm/rowSums(abs(tm))
ww<-wood_data|>mutate(window=(row_number()-1)%/%50)|>group_by(window)|>summarise(start=min(TIME),duration=max(TIME)-min(TIME),across(all_of(pitch_cols),mean),.groups="drop")
cm<-as.matrix(ww[,pitch_cols]);cn<-cm/(rowSums(abs(cm))+1e-10)
dm<-matrix(NA,nrow(cn),nrow(tm))
for(k in seq_len(nrow(tm))){d<-sweep(cn,2,tm[k,]);dm[,k]<-sqrt(rowSums(d^2))}
kl<-expand_grid(wi=seq_len(nrow(ww)),ki=seq_len(nrow(key_templates)))|>
  mutate(start=ww$start[wi],duration=ww$duration[wi],name=key_templates$name[ki],d=map2_dbl(wi,ki,\(w,k)dm[w,k]))|>
  mutate(name=factor(name,levels=key_templates$name))
ggplot(kl,aes(x=start+duration/2,width=duration,y=name,fill=d))+geom_tile()+
  scale_fill_viridis_c(guide="none")+theme_freq(11)+
  labs(title="Keygram: 'Wood'",subtitle="Krumhansl-Kessler key profile matching (darker = closer match)",x="Time (s)",y="")+
  theme(axis.text.y=element_text(size=7))
```

The keygram matches the chroma content of each time window against all 24 major and minor key templates using the Krumhansl-Kessler profiles. Darker horizontal bands indicate keys that consistently match well throughout the track. For *Wood*, the darkest bands correspond to Ab major and F minor, both of which remain stable across the entire duration with no visible modulation. This aligns with the key profile above and confirms that the track stays in a single tonal region throughout. The absence of any key change is consistent with the block structure observed in the Week 9 SSM and shows that compositional form in this genre relies on harmonic repetition rather than harmonic development. In traditional tonal music, key changes serve as a primary structural device to create tension and resolution, but EDM takes a fundamentally different approach: tension is created through textural buildup and release rather than harmonic motion.


