---
title: "Week 11: Tempo & Rhythm"
---

```{r setup}
#| include: false
source("setup.R")
```

## Temporal Structure and Rhythmic Organisation

Tempo is one of the most immediate perceptual differences between Minimal Tech House and Hard House. This section combines corpus-level analysis (tempo distribution) with track-level analysis (novelty functions and tempograms) to examine how rhythmic pacing and pulse stability differ between the two genres.

## Corpus Tempo Distribution

A histogram with 2 BPM bins compares the tempo distribution across both genres using overlaid bars. This approach reveals not just the central tendency but the full distributional shape, including overlap zones and bimodality.

```{r}
#| code-summary: "Histogram: tempo distribution per genre"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"

# Overlaid histogram with 2 BPM bins to compare tempo distributions
# position = "identity" overlays rather than stacks the bars
ggplot(corpus, aes(x = Tempo, fill = Genre)) +
  geom_histogram(binwidth = 2, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = pal) +
  theme_freq(12) +
  labs(
    title = "Tempo Distribution Across the Corpus",
    subtitle = "Each bar = 2 BPM | 861 tracks",
    x = "Tempo (BPM)", y = "Number of Tracks", fill = NULL
  )
```

The histogram confirms that tempo is one of the strongest differentiators between the two genres. Hard House clusters sharply around 138–150 BPM, reflecting the genre's aggressive, high-energy character. Minimal Tech House concentrates around 120–128 BPM, which is the classic house tempo range associated with groove-oriented, danceable rhythms. The two distributions barely overlap, with a near-empty zone between 128 and 138 BPM functioning as a natural boundary. Tempo alone could serve as a reasonably effective genre predictor for these two subgenres.

## Novelty Functions: First 60 Seconds

Novelty functions are computed by Sonic Visualiser and measure the spectral flux (rate of change) in the audio signal over time. Peaks correspond to rhythmic onsets. Two representative tracks are compared in a faceted line plot.

```{r}
#| code-summary: "Novelty functions: spectral flux over time"
#| fig-width: 10
#| fig-height: 5
#| out-width: "100%"

# Load novelty curves exported from Sonic Visualiser
# Columns: TIME (seconds), VALUE (spectral flux magnitude)
hard_nov <- read_csv("mytrack-novelty.csv", col_names = c("TIME", "VALUE"),
  show_col_types = FALSE) |> mutate(Genre = "Hard House: It Ain't Over (DJ Heartstring)")
min_nov <- read_csv("minimal-novelty.csv", col_names = c("TIME", "VALUE"),
  show_col_types = FALSE) |> mutate(Genre = "Minimal Tech House: Overflow (Kolter)")

# Faceted line plot comparing the first 60 seconds of each track
bind_rows(hard_nov, min_nov) |>
  ggplot(aes(x = TIME, y = VALUE, colour = Genre)) +
  geom_line(alpha = 0.8, linewidth = 0.4) +
  scale_colour_manual(values = c("Hard House: It Ain't Over (DJ Heartstring)" = "#ff2d55",
    "Minimal Tech House: Overflow (Kolter)" = "#00d4ff")) +
  facet_wrap(~Genre, ncol = 1, scales = "free_y") +
  xlim(0, 60) +
  theme_freq(11) +
  labs(
    title = "Novelty Functions: It Ain't Over vs Overflow",
    subtitle = "Onset strength over the first 60 seconds",
    x = "Time (s)", y = "Novelty", colour = NULL
  ) +
  theme(legend.position = "none")
```

The novelty functions capture how abruptly the spectral content changes over time, which closely tracks rhythmic onsets. DJ Heartstring's *It Ain't Over* shows a pattern of sharp, evenly-spaced peaks with high amplitude, a direct reflection of its relentless four-on-the-floor kick drum pattern where every beat hits hard. Kolter's *Overflow* shows a more irregular pattern with lower peak amplitude, consistent with its approach of gradual, texture-based development where rhythmic elements are layered subtly rather than delivered with force.

## Tempograms: It Ain't Over by DJ Heartstring (Hard House)

Tempograms visualise how rhythmic energy is distributed across different tempo frequencies over time. The ACT (autocorrelation) method detects periodicity in the onset signal, while the DFT (Fourier) method captures harmonic structure in the rhythm. Both are rendered as heatmaps using `geom_raster`.

::: {.panel-tabset}

### ACT
```{r}
#| code-summary: "ACT tempogram: Hard House"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"

# Load ACT tempogram data and reshape from wide to long format
# Reciprocal y-axis transforms lag values to BPM
read_csv("mytrack-act.csv", show_col_types = FALSE) |>
  pivot_longer(-TIME, names_to = "tempo") |>
  mutate(tempo = as.numeric(tempo)) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_y_continuous(transform = c("reciprocal", "reverse"), breaks = seq(50, 350, 50)) +
  scale_fill_viridis_c(guide = "none") +
  theme_freq(11) + theme(panel.grid = element_blank()) +
  labs(title = "ACT Tempogram: It Ain't Over (Hard House)", x = "Time (s)", y = "Tempo (BPM)")
```

### DFT
```{r}
#| code-summary: "DFT tempogram: Hard House"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"

# DFT tempogram captures harmonic structure in the rhythm
read_csv("mytrack-dft.csv", show_col_types = FALSE) |>
  pivot_longer(-TIME, names_to = "tempo") |>
  mutate(tempo = as.numeric(tempo)) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  theme_freq(11) + theme(panel.grid = element_blank()) +
  labs(title = "DFT Tempogram: It Ain't Over (Hard House)", x = "Time (s)", y = "Tempo (BPM)")
```
:::

## Tempograms: Overflow by Kolter (Minimal Tech House)

::: {.panel-tabset}

### ACT
```{r}
#| code-summary: "ACT tempogram: Minimal Tech House"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"
read_csv("minimal-act.csv",show_col_types=F)|>pivot_longer(-TIME,names_to="tempo")|>mutate(tempo=as.numeric(tempo))|>
  ggplot(aes(x=TIME,y=tempo,fill=value))+geom_raster()+scale_y_continuous(transform=c("reciprocal","reverse"),breaks=seq(50,350,50))+
  scale_fill_viridis_c(guide="none")+theme_freq(11)+theme(panel.grid=element_blank())+labs(title="ACT Tempogram: Overflow (Minimal Tech House)",x="Time (s)",y="Tempo (BPM)")
```

### DFT
```{r}
#| code-summary: "DFT tempogram: Minimal Tech House"
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"
read_csv("minimal-dft.csv",show_col_types=F)|>pivot_longer(-TIME,names_to="tempo")|>mutate(tempo=as.numeric(tempo))|>
  ggplot(aes(x=TIME,y=tempo,fill=value))+geom_raster()+scale_fill_viridis_c(guide="none")+
  theme_freq(11)+theme(panel.grid=element_blank())+labs(title="DFT Tempogram: Overflow (Minimal Tech House)",x="Time (s)",y="Tempo (BPM)")
```
:::

### Interpretation

The tempograms complement the novelty analysis by showing how rhythmic energy is distributed across tempo frequencies over time. The Hard House ACT tempogram displays a strong, consistent horizontal band at the track's tempo, pointing to a metrically rigid, unwavering beat. The DFT tempogram reveals additional energy at integer multiples and sub-multiples of the main tempo, harmonic artifacts that reflect the regularity of the pulse. For Minimal Tech House, the energy distribution is noticeably more diffuse: the rhythmic pulse is present but less forcefully articulated, consistent with a genre that favours subtlety over impact.

Together with the self-similarity analysis from Week 9, these tempograms complete the picture. Both genres are loop-based and repetitive, but Hard House asserts this through rhythmic rigidity, while Minimal Tech House expresses it through gradual textural continuity.
